Historical Soil Organic Carbon Budget

### 2020-07-01 SOCbudgetPaper/methods.Rmd
Update methods.Rmd from kristine first try of merging in davids changes

Following the tier 2 approach of the refinement of IPCC guidelines vol. 4 (@ipcc_2019_2019), we estimate soil organic carbon (SOC) stocks for cropland and all other land represented by natural vegetation on half-degree <!-- (55km at the equator) put this in for clarification? --> resolution from 1975 to 2010. We assume the actual SOC state converges towards a stable steady state, which itself is changing over time and space depending on biophysical, climatic and agronomic conditions. Therefore we conduct the following three steps within each yearly timestep: 
(1) We calculate annual land-use (sub-)type-specific steady states and decay rates for SOC stocks, 
(2) We account for land conversion by transferring SOC between land-use types and 
(3) We estimate SOC stocks based on the stocks of the previous time period, the steady state stocks and the decay rate.

<!-- maybe make it more clear what the steady state is ? what does actual (actual SOC) mean here ? what does actual mean?  quantity of total carbon? so far not so clear to me, and its a bit confusing when you say its still changing over time.  i think these lines need to be more clear, something like "which still includes changing internal dynamics over space and time depending on biophysical..." -->

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{eq}$ are given by 
\begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
\label{eq:inoutflow}
\end{equation}
with $C^{\textrm{in}}$ being carbon inputs to the soil, $k$ denoting the soil organic carbon decay rate; as well as $i$ representing grid cell indices and $t$ years. We use for our calculations the steady-state method from the refinement of the IPCC guidelines vol. 4 (@ipcc_2019_2019) for mineral soils, which assumes three soil carbon sub-pools (active, slow and passive) and interactions between them. Annual carbon inflow to each sub-pool (see \@ref(sec:carboninputs)) and annual decay rates (see \@ref(sec:tier2)) of each sub-pool are the key components to determining steady-state SOC stocks.

@@

We account for different carbon input sources depending on the two land-use types we distinguish, croplands and natural vegetation (see table \@ref(tab:datasourceinputs)). Carbon sources for cropland are recycled crop residues, below ground biomass of crops (for both see \@ref(sec:residues)) and recycled manure (see \ref(sec:livst_manure)); For natural vegetation litterfall (\ref(@LPJmL4_1)) litterfall is the (main/only) source of carbon.

<!-- i took out in particular because this means there are more than the 3 you list? you could say "C sources for cropland are namely..." instead of in particular, but its a bit redundant here  -->

@@

The sub-pool specific decay rates are influenced by climatic conditions, biophysical and biochemical soil properties as well as management factors that all vary over time (t) and space (i). Following the steady-state method of the refinement of the IPCC guidelines vol. 4 (@ipcc_2019_2019) for mineral soils <!-- Can you shorten this because you already said it word-for-word in previous paragraph? --> we consider temperature (temp), water (wat), sand fraction (sf) and tillage (till) effects to account for spatial variation of decay rates. Thus $k_{sub}$ is given by

@@

<!-- do u test the assumption of 15 year spin up not influencing results? I feel like this si something people may critique -->

@@

We compile country-specific FAO production and cropland statistics ([@FAOSTAT]) to a comprehensive and consistent data suite. <!-- I don't know what you are trying to say with "...to a comprehensive and consistent data suite"    into a harmonized and consistent data set? comprehensive in what way? the term data suite sounds very odd to me --> The data is prepared in 5 year time steps from 1965 to 2010, which, together with the spin up phase, restricts our analysis to time span from 1975 to 2010. For all the following data, if not declared differently, we interpolate values linearly between the time steps.

@@

##### Residue Usage
We compile out of our 17 crop groups, three residue groups (straw, high-lignin and low-lignin residues) with additional demand for other purposes and one residue with no double use (see \@ref(append:Table_kcr2kres)). Residue feed demand for five different livestock groups is based on country- and residue-group-specific feed baskets (see [@weindl]), taking available AGR biomass as well as livestock productivity into account. We estimate, for low-income countries according to World Bank definitions ($<\,1000\,\tfrac{USD}{yr}$) <!-- specify per capita -->,  a material-use share for the straw residues group <!-- is straw residues a group? --> of 5% and a fuel-share of 10% for all used residues groups <!-- does this mean "10%  of all residues used as fuel in low income.." ?--> in low income countries according to worldbank definitions ($<\,1000\,\tfrac{USD}{yr}$). For high-income countries, ($>\,10000\,\tfrac{USD}{yr}$ no withdrawal for material or fuel use is assumend, and middle-income countries use shares are linearly interpolated based on per-capita income. The remaining AGR as well as all BGR are assumend to be recycled to the soil. We limit high recycling shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.

@@

#### Livestock Distribution
To disaggregate country level FAOSTAT livestock production values to the half-degree scale, we use the following rule-based assumptions, inspired by the approach of [@gilbert], and uses feed basket assumptions based on a revised methodology from @weindl. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems. <!-- "inspired by" sounds vague to me and personally I would never write it. "loosely based on" or "drawing from" is what I would say... -->
For ruminants, we assume that livestock is located where the production of feed occurs. We distingush between grazed pasture, which is converted into livestock products from extensive systems; and all other (crop-based) feeds, which we consider to be consumend in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to divide between intensive and extensive production systems. For low-income countries according to worldbank definitions (<1000 USD/yr), we assume extensive production systems. We locate them according to the share of built-up areas based on the idea that these animals are held in households, subsistence or small-holder farming systems with a high labour per animal ratio <!-- after "based on" it is not clear to me. THe animals are held in EITHER households, OR subsistence and small holders? which of these has the high labour per animal ratio? --> Intensive production is distributed within a country using again share of crop-based feed production,  assuming that feed availability is the most driving factor for livestock location. <!-- what happens with extensive/intensive in higher income countries is missing or unclear to me -->

#### Manure Excretion, Storage and Recycling
Manure production and management is based on a revised methodology of ([@bodirsky2012]) and is presented here due to its central role in soil carbon modelling. Based on the gridded livestock distribution we calculate excretions by estimating the nitrogen balance of the livestock system on the basis of comprehensive livestock feed baskets ([@weindl]), assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios (given by [(@ipcc_2019_2019)]). 
Depending on the feed system we assume manure to be handled in four different ways:
All manure orginated from pasture feed intake is excreted directly to pastures and rangelands (pasture grazing), deducting manure collected as fuel. Manure fuel shares are estimated using IPCC default values (@ippc_2006_2006). 
Whereas for low-income countries according to worldbank definitions (<1000 USD/yr) <!-- I feel like you don't ahve to say "according to world bank... " after the first time -->, we adopt a share of 25% of crop residues in feed intake directly consumend and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries ($>\,10000\,\tfrac{USD}{yr}$; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually recycled to the soil, we account for carbon losses during storage and recycling shares in different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not recycled. For manure stored in different animal waste management systems we compiled carbon loss rates partly depending on the nitrogen loss rates as specified in [@bodirsky2012] (see \@ref(append:Table_clossAWMS))). We limit high application shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.

>>>>

Following the tier 2 approach of the refinement of IPCC guidelines vol. 4 (@ipcc_2019_2019; short Tier 2 steady-state method) , we estimate soil organic carbon (SOC) stocks for cropland and all other land represented by natural vegetation on half-degree resolution from 1975 to 2010. We assume the current SOC state converges towards a steady state, which itself is depending on biophysical, climatic and agronomic conditions. Therefore we conduct the following three steps within each yearly timestep: 
(1) We calculate annual land-use (sub-)type-specific steady states and decay rates for SOC stocks, 
(2) We account for land conversion by transferring SOC between land-use types and 
(3) We estimate SOC stocks based on the stocks of the previous time period, the steady state stocks and the decay rate.

@@

In a simple first order kinetic approach the steady-state soil organic carbon stocks $SOC^{eq}$ are given by 
\begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
\label{eq:inoutflow}
\end{equation}
with $C^{\textrm{in}}$ being carbon inputs to the soil, $k$ denoting the soil organic carbon decay rate; as well as $i$ representing grid cell indices and $t$ years. We use for our calculations the Tier 2 steady-state method, which assumes three soil carbon sub-pools (active, slow and passive) and interactions between them. Annual carbon inflow to each sub-pool (see \@ref(sec:carboninputs)) and annual decay rates (see \@ref(sec:tier2)) of each sub-pool are the key components to determining steady-state SOC stocks.

@@

We account for different carbon input sources depending on the two land-use types we distinguish, croplands and natural vegetation (see table \@ref(tab:datasourceinputs)). Carbon sources for cropland are recycled crop residues, below ground biomass of crops (for both see \@ref(sec:residues)) and recycled manure (see \ref(sec:livst_manure)); for natural vegetation litterfall (\ref(@LPJmL4_1)) litterfall is the only source of carbon.

@@

The sub-pool specific decay rates are influenced by climatic conditions, biophysical and biochemical soil properties as well as management factors that all vary over time (t) and space (i). Following the Tier 2 steady-state method we consider temperature (temp), water (wat), sand fraction (sf) and tillage (till) effects to account for spatial variation of decay rates. Thus $k_{sub}$ is given by

@@

@@

We compile country-specific FAO production and cropland statistics ([@FAOSTAT]) to a harmonized and consistent data set. The data is prepared in 5 year time steps from 1965 to 2010, which, together with the spin up phase, restricts our analysis to time span from 1975 to 2010. For all the following data, if not declared differently, we interpolate values linearly between the time steps and hold them constant before 1965.

@@

##### Residue Usage
We compile out of our 17 crop groups, three residue groups (straw, high-lignin and low-lignin residues) with additional demand for other purposes and one residue with no double use (see \@ref(append:Table_kcr2kres)). Residue feed demand for five different livestock groups is based on country- and residue-group-specific feed baskets (see [@weindl]), taking available AGR biomass as well as livestock productivity into account. We estimate, for low-income countries ($<\,1000\,\tfrac{USD}{yr}$) per capita,  a material-use share for straw residues of 5% and a fuel-share of 10% for all used residues groups in low income countries ($<\,1000\,\tfrac{USD}{yr}$). For high-income countries, ($>\,10000\,\tfrac{USD}{yr}$ no withdrawal for material or fuel use is assumend, and middle-income countries use shares are linearly interpolated based on per-capita income. The remaining AGR as well as all BGR are assumend to be recycled to the soil. We limit high recycling shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.

@@

#### Livestock Distribution
To disaggregate country level FAOSTAT livestock production values to the half-degree scale, we use the following rule-based assumptions, drawing from by the approach of [@gilbert], and uses feed basket assumptions based on a revised methodology from @weindl. We differentiate between ruminant and monogastric systems, as well as extensive and intensive systems.
For ruminants, we assume that livestock is located where the production of feed occurs. We distingush between grazed pasture, which is converted into livestock products from extensive systems; and all other (crop-based) feeds, which we consider to be consumend in intensive systems.
For poultry, egg and monogastric meat production we use the per-capita income of the country to divide between intensive and extensive production systems. For low-income countries (<1000 USD/yr), we assume extensive production systems. We locate them according to the share of built-up areas based on the idea that these animals are held in subsistence or small-holder farming systems with a high labour per animal ratio. Intensive production associated with higher income countries, is distributed within a country using again share of crop-based feed production,  assuming that feed availability is the most driving factor for livestock location.

#### Manure Excretion, Storage and Recycling
Manure production and management is based on a revised methodology of ([@bodirsky2012]) and is presented here due to its central role in soil carbon modelling. Based on the gridded livestock distribution we calculate excretions by estimating the nitrogen balance of the livestock system on the basis of comprehensive livestock feed baskets ([@weindl]), assuming that all nitrogen in protein feed intake, minus the nitrogen in the slaughter mass, is excreted. Carbon in excreted manure is estimated by applying fixed C:N ratios (given by [(@ipcc_2019_2019)]). 
Depending on the feed system we assume manure to be handled in four different ways:
All manure orginated from pasture feed intake is excreted directly to pastures and rangelands (pasture grazing), deducting manure collected as fuel. Manure fuel shares are estimated using IPCC default values (@ippc_2006_2006). 
Whereas for low-income countries (<1000 USD/yr), we adopt a share of 25% of crop residues in feed intake directly consumend and excreted on crop fields (stubble grazing), we do not consider any stubble grazing in high-income countries ($>\,10000\,\tfrac{USD}{yr}$; middle-income countries see linearly interpolated shares depending on their per-capita income.
For all other feed items we assume the manure to be stored in animal waste management systems associated with livestock housing. 
To estimate the carbon actually recycled to the soil, we account for carbon losses during storage and recycling shares in different animal waste management and grazing systems. Whereas we assume no losses for pasture and stubble grazing, we consider that the manure collected as fuel is not recycled. For manure stored in different animal waste management systems we compiled carbon loss rates partly depending on the nitrogen loss rates as specified in [@bodirsky2012] (see \@ref(append:Table_clossAWMS))). We limit high application shares at $10\tfrac{\unit{tC}}{\unit{ha}}$ to correct for outliers.
